\documentclass{beamer}

\usepackage{fontspec}
\usepackage{polyglossia}

\usepackage{graphicx}
\usepackage{color}
\usepackage{url}

\makeatletter
\newcommand\listofframes{\@starttoc{lbf}}
\makeatother

\addtobeamertemplate{frametitle}{}{%
  \addcontentsline{lbf}{section}{\protect\makebox[2em][l]{%
    \protect\usebeamercolor[fg]{structure}\insertframenumber\hfill}%
  \insertframetitle\par}%
}


\title{FIWCLUL Conclusion / Business Meeting / Future Plans\\
\scriptsize{in Tromsø, 2015}}
\author{Tommi A Pirinen \scriptsize \guilsinglleft tommi.pirinen@computing.dcu.ie \guilsinglright }
\institute{DCU, CNGL}
\date{\today}

\begin{document}

\selectlanguage{english}

\maketitle

\begin{frame}
    \frametitle{Agenda}
    \listofframes
\end{frame}

\begin{frame}
    \frametitle{Organising meeting}
    \begin{itemize}
        \item Free beer
        \item Decide how to organise meeting
        \item Modify agenda
        \item Stuff
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Overview of this year}
    \begin{itemize}
        \item 14 papers, 12 accepted
        \item variety of topics: Lexicography, OCR, Web-based interfaces,
            \ldots
        \item potential co-operations, interactions at the workshop,
            maybe methods and platforms to concentrate efforts
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Next events}
    \begin{itemize}
        \item (FU conference is in Oulu in Autumn)
        \item future IWCLULs: yearly or less? co-located with bigger
            conferences?
        \item Suggested: Veroika et al. organise in Hungary, possibly
            co-locate with magyar számítógépes nyelvészeti konferencia XI.?
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Organising ourselves}
    \begin{itemize}
        \item Mailing list? (We already have a bunch unused...)
        \item would ACL SIG be a good way to organise efforts?
            \begin{itemize}
                \item possibility to have workshops at ACL conferences,
                    with some financial backing from ACL
                \item requires bureaucracy: a board, an org, members,
                    ACL business meetings etc.
            \end{itemize}
        \item N.B. budget this year was ad hoc made up with good help
            of giellatekno and uit
        \item other options?
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Any Other Business (AOB)}
    \begin{itemize}
        \item open floor
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Adjourn---Social Event}
    \begin{itemize}
        \item We will meet in the pub in couple of hours, the social
            event is optional but recommended.
    \end{itemize}
\end{frame}

\end{document}
